\documentclass[dvipsnames,11pt]{article}

\input{SimpleStef}
\usetikzlibrary{positioning,arrows.meta,backgrounds,calc}

\graphicspath{ {./images/} }
% LAST IMAGE UPDATE: 20-02-2026, 01:02 AM

\setlength{\headheight}{25pt} % Non so perché ma senza questo da un piccolo errore


\pagestyle{fancy}
\fancyhf{} 
\fancyhead[L]{\textsc{\small{20285\\ Advanced Macroeconomics}}} 
\fancyhead[R]{\textsc{\small Lo Schiavo et al.}} 
\fancyhead[C]{\textbf{Problem set 4}}
\fancyfoot[C]{\thepage} 
\renewcommand{\headrulewidth}{0.4pt} 
\renewcommand{\footrulewidth}{0pt}

\title{20285 Advanced Macroeconomics \\[0.5cm] \Huge \textsc{Problem set 4}}

\author{
        Laura Lo Schiavo \\
        \texttt{3317819}
        \and 
        Trace Miller \\
        \texttt{3319805}
        \and 
        Pietro Sferruzza \\
        \texttt{3323719}
        \and
        Stefano Graziosi\\
        \texttt{3174891}
        }

\date{\emph{Compiled on \today}}

\begin{document}

\maketitle

\tableofcontents

\section{Optimal price setting in the Taylor model} 

    In this exercise, we explore the Taylor model of staggered price setting, where each firm resets its price for \( N \) periods. In each period, a fraction \( \frac{1}{N} \) of firms reset their prices, while the remaining firms keep their prices fixed. 
    Each firm \( i \) produces a differentiated good with technology:
    \[
    Y_t(i) = A_t N_t(i),
    \]
    where \( A_t \) represents the level of technology, evolving exogenously as \( a_t = \log A_t \) with a stationary stochastic process. A firm resetting its price \( P_t^* \) in period \( t \) seeks to maximize its expected profits over \( N \) periods:
    
    \[
    \max_{P_t^*} \sum_{k=0}^{N-1} \mathbb{E}_t \left\{ Q_{t,t+k} \left( \frac{1}{P_{t+k}} \right) \left( P_t^* Y_{t+k|t} - \Psi_{t+k}(Y_{t+k|t}) \right) \right\}
    \]
    subject to:
    \[
    Y_{t+k|t} = \left( \frac{P_t^*}{P_{t+k}} \right)^{-\epsilon} C_{t+k},
    \]
    where \( Q_{t,t+k} = \beta^k \left( \frac{C_{t+k}}{C_t} \right)^{-\sigma} \) is the stochastic discount factor, $ \Psi_{t+k}(Y_{t+k|t})$ are the nominal cost, and \( C_{t+k} \) is aggregate consumption in period \( t+k \).
    
    \begin{enumerate}
        \item Show that the optimal reset price \( P_t^* \) must satisfy the first-order condition:
        \[
        \sum_{k=0}^{N-1} \mathbb{E}_t \left\{ \left( \frac{1}{P_{t+k}} \right) Q_{t,t+k} Y_{t+k|t} \left( P_t^* - \mathcal{M} \psi_{t+k} \right) \right\} = 0,
        \]
        where \( \psi_{t+k} = \Psi'(\cdot) \) is the nominal marginal cost and \( \mathcal{M} = \frac{\epsilon}{\epsilon - 1} \).
    
        \textit{[Hint: Set up the first-order condition by differentiating the profit function with respect to \( P_t^* \). Remember to substitute the demand function and use the stochastic discount factor for nominal payoffs.]}
    
            \begin{solution}
    
                
                
            \end{solution}
    
        \item Derive the log-linearized version of the optimal price-setting rule around a zero-inflation steady state:
        \[
        p_t^* = \mu + \sum_{k=0}^{N-1} \omega_k \mathbb{E}_t \left\{ \varphi_{t+k} \right\},
        \]
        where \( \omega_k = \frac{\beta^k (1 - \beta)}{1 - \beta^N} \), \( \varphi_{t+k} = \log \psi_{t+k} \), and \( \mu = \log \mathcal{M} \).
    
        Interpret this condition.
    
        \textit{[Hint: Approximate the optimal price around a zero-inflation steady state by log-linearizing the price-setting rule. This will involve taking logarithms and expectations.]}
    
            \begin{solution}
    
                
                
            \end{solution}
    
        \item How does the Taylor model’s price-setting rule differ from the one generated by the Calvo model? Discuss the economic intuition behind these differences, focusing on the implications for inflation persistence. How do they compare in terms of tractability? 
    
            \begin{solution}
    
                
                
            \end{solution}
    
        \item Recalling that the aggregate price index \( P_t \) is defined by:
        \[
        P_t = \left( \int_0^1 P_t(i)^{1 - \epsilon} \, di \right)^{\frac{1}{1 - \epsilon}},
        \]
        show that, around a zero-inflation steady state, the (log) price level will satisfy:
        \[
        p_t = \frac{1}{N} \sum_{k=0}^{N-1} p_{t-k}^*.
        \]
    
        \textit{[Hint: Argue why we have symmetric pricing for firms within each cohort and derive the relationship between the log price level \( p_t \) and the optimal reset prices \( p_{t-k}^* \).]}
    
            \begin{solution}
    
                
                
            \end{solution}
    
    \end{enumerate}

\section{The Phillips Curve with Price Indexation}

    In the New Keynesian model, there is a probability $\theta$ that a firm will not be able to change its nominal price in a given time period. Firms that do have an opportunity to change price at time $t$ choose a reset price $p^*_t$ according to:
    \[
    p^*_t = (1 - \beta \theta) \sum_{\ell=0}^\infty (\beta \theta)^\ell \mathbb{E}_t [p_{t+\ell} + \gamma \widehat{mc}_{t + \ell }],
    \]
    where $\beta$ is the discount factor, $p_t$ is the price level, $\widehat{mc}_t$ is average marginal cost, and $\gamma = \frac{1 - \alpha}{1 - \alpha + \alpha \epsilon}$ is a positive coefficient. All variables are given as log deviations from their steady-state values. The general price level $p_t$ is determined by:
    \[
    p_t = (1 - \theta) \sum_{\ell=0}^\infty \theta^\ell p^*_{t-\ell}.
    \]

    \begin{enumerate}
    
        \item Provide an intuitive explanation of the two equations above.
    
            \begin{solution}
    
                
                
            \end{solution}
    
        \item Show that the reset-price and price-level equations are equivalent to:
        \[
        p^*_t = \beta \theta \mathbb{E}_t p^*_{t+1} + (1 - \beta \theta)(p_t +  \gamma \widehat{mc}_{t}),
        \]
        and
        \[
        p_t = \theta p_{t-1} + (1 - \theta) p^*_t.
        \]
    
            \begin{solution}
    
                
                
            \end{solution}
    
        \item By combining the two equations, derive the New Keynesian Phillips curve for inflation $\pi_t = p_t - p_{t-1}$:
        \[
        \pi_t = \beta \mathbb{E}_t \pi_{t+1} + \kappa \widetilde{y}_t,
        \]
        using that $\widehat{mc}_t = \left( \sigma + \frac{\alpha + \phi}{1 - \alpha}  \right) \widetilde{y}_t$, and where $\kappa = \gamma \left( \sigma + \frac{\alpha + \phi}{1 - \alpha}  \right) \left( \frac{(1 - \theta)(1 - \beta \theta)}{\theta}    \right)$
    
            \begin{solution}
    
                
                
            \end{solution}
            
    \end{enumerate}

    \noindent Next suppose that when a firm does not have an opportunity to choose a new reset price optimally at time $t$ (probability $\theta$), the firm adjusts its old nominal price in line with any past inflation $\pi_{t-1}$ that has occurred.\footnote{For a more detailed treatment, see for example "Trend Inflation, Indexation, and Inflation Persistence in the New Keynesian Phillips Curve" by Cogley and Sbordone (\href{https://www.aeaweb.org/articles?id=10.1257/aer.98.5.2101}{AER 2008}).} This means the equation for the price level is now:
    \[
    p_t = \theta (p_{t-1} + \pi_{t-1}) + (1 - \theta) p^*_t.
    \]
    
    \noindent When choosing the reset price $p^*_t$, firms take account of the indexation of prices to past inflation. Hence the new equation for the reset price is:
    \[
    p^*_t = (1 - \beta \theta) \sum_{\ell=0}^\infty (\beta \theta)^\ell \mathbb{E}_t [p_{t+\ell} - (p_{t+\ell-1} - p_{t-1}) +  \gamma \widehat{mc}_{t+\ell}].
    \]

    \begin{enumerate}[resume]

        \item Explain the intuition for the new reset-price equation.

            \begin{solution}

            
            
            \end{solution}

        \item Derive the new Phillips curve. Comment on any differences relative to the standard New Keynesian Phillips curve. 

            \textit{[Hint: Start by putting the reset-price equation in a simpler recursive form.]} 

            \begin{solution}

            
            
        \end{solution}

    \end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




% \include{2025/PS/PS 4/ex2_sol}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{The Phillips Curve}

    \subsection{The Classic Phillips Curve}

        For this exercise, download the following series at quarterly frequency from 1960:Q1 through 2025:Q4 from \href{https://fred.stlouisfed.org}{FRED}:
        \begin{itemize}
            \item $u_t$ = \textit{Unemployment Rate} (Unit: Percent, Seasonally Adjusted);
            
            \item $\pi_t$ = \textit{Personal Consumption Expenditures: Chain-type Price Index} (Unit: Percent Change from Year Ago, Seasonally Adjusted);
        \end{itemize}
        
        \begin{enumerate}
            \item Construct a series for expected price inflation, measured as $\mathbb{\widetilde{E}}_{t}[\pi_{t+1}] = \frac{1}{4}(\pi_{t-1} + \pi_{t-2} + \pi_{t-3} + \pi_{t-4})$. Briefly interpret and comment on this measure of inflation expectations.  

                \begin{solution}

                    
                    
                \end{solution}
            
            \item Create a scatterplot of \(u_t\) (x-axis) and \( (\pi_t - \mathbb{\widetilde{E}}_{t}[\pi_{t+1}] ) \) (y-axis) for the entire period. Fit a simple OLS regression line and plot it in the same graph. Briefly describe your findings. What is \( (\pi_t - \mathbb{\widetilde{E}}_{t}[\pi_{t+1}] ) \) measuring? 
            \item Repeat the exercise from above for the following subperiod:
            \begin{itemize}
                \item[a.] 1960:Q1 -- 1986:Q4 (blue)
                \item[b.] 1987:Q1 -- 2007:Q4 (green)
                \item[c.] 2008:Q1 -- 2019:Q4 (red)
                \item[d.] 2020:Q1 -- 2025:Q4 (black)
            \end{itemize}
            Compare the scatterplots and slopes  over time and discuss.
    
                \begin{solution}

                    
                    
                \end{solution}
            
        \end{enumerate}

    \subsection{Household expectations}
    
        \begin{enumerate}[resume]
            \item Read and summarize the key question and findings of "Is the Phillips Curve Alive and Well after All? Inflation Expectations and the Missing Disinflation" by Coibion and Gorodnichenko (\href{https://www.aeaweb.org/articles?id=10.1257/mac.20130306}{AEJ: Macro 2015}). 
        \end{enumerate}

        \noindent In this exercise, we focus on the third potential explanation for the missing disinflation discussed in Section IV of the paper. Begin by downloading the following time series: 
        \begin{itemize}
            \item[\(\Rightarrow\)] $\mathbb{E}^{MSC}_t[\pi_{t+4}]$ = \textit{Household inflation forecast, Median 1-year ahead} (Source: \href{https://data.sca.isr.umich.edu/data-archive/mine.php}{University of Michigan - Survey of Consumers}, Table 32, time span: 1978:Q1-2025:Q4).
        \end{itemize}

        \noindent Next, we will estimate the Phillips curve using household forecasts. Specifically, we consider a Friedman (1968)-type expectations-augmented Phillips curve:
        \begin{equation}\label{eq1} 
            \pi_t - \mathbb{E}_t[\pi_{t+1}] = c  + \kappa u_t + v_t,
        \end{equation}
        as in equation (1) of Coibion and Gorodnichenko (2015). We are interested in estimating $\kappa$, the slope of the Phillips curve. To do so, we will implement different measures of inflation expectations, $\mathbb{E}_t[\pi_{t+1}]$.
        
        \begin{enumerate}[resume]
            \item First, plot, describe and compare the time series of the two measures of inflation expectations, $\mathbb{\widetilde{E}}_t[\pi_{t+1}]$ and $\mathbb{E}^{MSC}_t[\pi_{t+4}]$). Why is it useful to use the median (instead of the mean for example) for household inflation expectations?
        
                \begin{solution}
        
                    
                    
                \end{solution}
            
            \item Next, estimate ($\ref{eq1}$) assuming that $\mathbb{E}_t[\pi_{t+1}] = \mathbb{E}^{MSC}_t[\pi_{t+4}]$. Create a scatter plot and include your estimated regression line, as in Figure 6 Panel B of the paper.
            
                Relate your findings to the results of Coibion and Gorodnichenko (2015) and discuss.  
        
                \begin{solution}
        
                
                    
                \end{solution}
        
             \item Briefly discuss the OLS estimation of the slope of the Phillips curves in 3.2, 3.3. and 3.6. 
        
                \textit{[Hint: Specifically, consider the identification problem of simultaneity bias in the given case. You may want to have a look at Section 2 and, more importantly, Section 3 of \href{https://www.journals.uchicago.edu/doi/full/10.1086/707181}{McLeay and Tenreyro (2019)}.]}
        
                \begin{solution}
        
                    
                    
                \end{solution}
                
        \end{enumerate}

    \subsection{Non-linear Phillips curves}

        \begin{enumerate}[resume]
            \item Read “The Slanted-L Phillips Curve” by Benigno and Eggertsson (2024) (\href{https://doi.org/10.1257/pandp.20241051}{AEA Papers and Proceedings, 2024}). Briefly summarize the authors’ key arguments.
        
            \item Replicate Figure (1) of the paper for the US. To do so, simply perform the following steps: 
    
            \begin{enumerate}
                \item Download quarterly data from the following series from Q1 2000 through Q4 2025 from \href{https://fred.stlouisfed.org}{FRED}:
                \begin{itemize}
                    \item $V_t$ = \textit{Job Openings: Total Non-Farm} (Unit: Level in Thousands, Seasonally Adjusted);
                    \item $U_t$ = \textit{Unemployment Level} (Unit: Thousands of Persons, Seasonally Adjusted);
                \end{itemize}
                
                \item For each period, compute the $V_t/U_t$ ratio. Finally, create a scatter plot of \(V_t/U_t\) (x-axis) and \(\pi_{t}\) (y-axis). 
        
                \item Briefly describe your findings and discuss.\footnote{For a more detailed descriptive background see \href{https://www.nber.org/papers/w31197}{Benigno and Eggertson(2024)} and \href{https://www.nber.org/papers/w33095}{Benigno and Eggertson(2024)}.} 
            \end{enumerate}

            \item Lastly, read Giulia Gitti's Job Market Paper (\href{https://drive.google.com/file/d/1ZD9WUJ06xZFO8bhZsQyy3tZZfk2yC8oH/view}{JMP, 2024}). Explain her approach of addressing potential endogeneity when relating labor market tightness and inflation. How does this relate to your findings in 3.8 and 3.9 ?
        \end{enumerate}


    

\printbibliography

\end{document}
